---
title: "complex-maze"
author: "Allison Davis"
date: "February 15, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data collection

```{r, warning=FALSE, message=FALSE}
library(curl)

raw <- curl("https://raw.githubusercontent.com/allisondavis/complex_maze/master/adavis_complex-complete.csv")

raw <- read.csv(raw, header = TRUE, sep = ",", stringsAsFactors = TRUE)

head(raw)

```


## Analyses {.tabset .tabset-fade .tabset-pills}

###  First look

I will first look at the data via boxplots for each trial vs species, as well as bar charts per species

```{r, warning=FALSE, message=FALSE}
library(ggplot2)

ggplot(raw, aes(x=SPP, y=T1.TIME, fill=SPP)) + geom_boxplot() + xlab("Species") + ylab("Time to completion (s)\n") + ggtitle("Trial 1\n") + geom_jitter(shape=16, position=position_jitter(0.2)) + theme_classic() + scale_fill_manual(values = c("#9C964A", "#FAD77B")) + theme(text = element_text(size = 16)) + theme(legend.position = "none")

ggplot(raw, aes(x=SPP, y=T2.TIME, fill=SPP)) + geom_boxplot() + xlab("Species") + ylab("Time to completion (s)\n") + ggtitle("Trial 2\n") + geom_jitter(shape=16, position=position_jitter(0.2)) + theme_classic() + scale_fill_manual(values = c("#9C964A", "#FAD77B")) + theme(text = element_text(size = 16)) + theme(legend.position = "none")

ggplot(raw, aes(x=SPP, y=T3.TIME, fill=SPP)) + geom_boxplot() + xlab("Species") + ylab("Time to completion (s)\n") + ggtitle("Trial 3\n") + geom_jitter(shape=16, position=position_jitter(0.2)) + theme_classic() + scale_fill_manual(values = c("#9C964A", "#FAD77B")) + theme(text = element_text(size = 16)) + theme(legend.position = "none")

#come back and get bar chart/boxplot/scatterplot to compare three trials within each spp

lat <- raw[raw$SPP == "lat",]
form <- raw[raw$SPP == "form",]

trial.er.lat <- lat[c(4,6,8)]
trial.tm.lat <- lat[9:11]

trial.er.form <- form[c(4,6,8)]
trial.tm.form <- form[c(9:11)]

par(mfrow=c(2,2))
boxplot(trial.er.lat, main="Errors for sailfin")
boxplot(trial.er.form, main="Errors for amazon")
boxplot(trial.tm.lat, main="Duration for sailfin")
boxplot(trial.tm.form, main ="Duration for amazon")


```


### Simple stats

* using t-test to detect differences in means between species in same trial, and between trials in same species

```{r}
#testing differences between species w/in same trial
t.test(lat$T1.I, form$T1.I, alternative = "two.sided")

t.test(lat$T2.I, form$T2.I, alternative = "two.sided")

t.test(lat$T3.I, form$T3.I, alternative = "two.sided")

#testing differences between trials w/in same species
t.test(lat$T1.I, lat$T2.I, alternative = "two.sided")

t.test(lat$T1.I, lat$T3.I, alternative = "two.sided")

t.test(lat$T2.I, lat$T3.I, alternative = "two.sided")


t.test(form$T1.I, form$T2.I, alternative = "two.sided")

t.test(form$T1.I, form$T3.I, alternative = "two.sided")

t.test(form$T2.I, form$T3.I, alternative = "two.sided")

```

<font color = "deeporchid4"> No significant differences found.</font>

* using F-test to detect different variations in differnt species in same trial, and different trials in same species

```{r}


#testing variances between species w/in same trial
var.test(lat$T1.I, form$T1.I, alternative = "two.sided")

var.test(lat$T2.I, form$T2.I, alternative = "two.sided")

var.test(lat$T3.I, form$T3.I, alternative = "two.sided")

#testing variances between trials w/in same species
var.test(lat$T1.I, lat$T2.I, alternative = "two.sided")

var.test(lat$T1.I, lat$T3.I, alternative = "two.sided")

var.test(lat$T2.I, lat$T3.I, alternative = "two.sided")


var.test(form$T1.I, form$T2.I, alternative = "two.sided")

var.test(form$T1.I, form$T3.I, alternative = "two.sided")

var.test(form$T2.I, form$T3.I, alternative = "two.sided")


```

* ***will repeat for time of trials***

* using t-test to detect differences in means between species in same trial, and between trials in same species

```{r}
#testing differences between species w/in same trial
t.test(lat$T1.TIME, form$T1.TIME, alternative = "two.sided")

t.test(lat$T2.TIME, form$T2.TIME, alternative = "two.sided")

t.test(lat$T3.TIME, form$T3.TIME, alternative = "two.sided")

#testing differences between trials w/in same species
t.test(lat$T1.TIME, lat$T2.TIME, alternative = "two.sided")

t.test(lat$T1.TIME, lat$T3.TIME, alternative = "two.sided")

t.test(lat$T2.TIME, lat$T3.TIME, alternative = "two.sided")


t.test(form$T1.TIME, form$T2.TIME, alternative = "two.sided")

t.test(form$T1.TIME, form$T3.TIME, alternative = "two.sided")

t.test(form$T2.TIME, form$T3.TIME, alternative = "two.sided")

```

<font color = "deeporchid4"> No significant differences found.</font>

* using F-test to detect different variations in differnt species in same trial, and different trials in same species

```{r}


#testing variances between species w/in same trial
var.test(lat$T1.TIME, form$T1.TIME, alternative = "two.sided")

var.test(lat$T2.TIME, form$T2.TIME, alternative = "two.sided")

var.test(lat$T3.TIME, form$T3.TIME, alternative = "two.sided")

#testing variances between trials w/in same species
var.test(lat$T1.TIME, lat$T2.TIME, alternative = "two.sided")

var.test(lat$T1.TIME, lat$T3.TIME, alternative = "two.sided")

var.test(lat$T2.TIME, lat$T3.TIME, alternative = "two.sided")


var.test(form$T1.TIME, form$T2.TIME, alternative = "two.sided")

var.test(form$T1.TIME, form$T3.TIME, alternative = "two.sided")

var.test(form$T2.TIME, form$T3.TIME, alternative = "two.sided")


```


<font color="deeporchid4"> Only significant difference is between trials 1 and 3 in amazons.</font>


###  Complex stats {.tabset .tabset-fade .tabset-pills}

#### **First try** 

Here I will use a generalized linear mixed-effects model to assess differences in the number of errors made across trials between the species.
  * species as fixed effect, individual ID as random effect, and trial number as covariate

Had to rearrange the data, so that each trial for each individual was on a separate line. Will upload now.

```{r, warning=FALSE, message=FALSE}
library(curl)

Mraw <- curl("https://raw.githubusercontent.com/allisondavis/complex_maze/master/mixed.format_complex-raw.csv")

Mraw <- read.csv(Mraw, header = TRUE, sep = ",", stringsAsFactors = TRUE)

head(Mraw)

```

Now let's check how the data are distributed. Normal distributions can use linear mixed models, whereas non-normal distributions can use generalized linear mixed models.

  
```{r, warning=FALSE, message=FALSE}
library(car)

qqp(Mraw$INC, "norm")  #fits normal fine, so can used LMM
qqp(Mraw$TIME, "norm") #fits ok, little off in top right; can try LMM but may want to try a different distribution
qqp(Mraw$TIME, "lnorm") #fits better, since no points are outside dashed lines; try out GLMM

boxplot(data=Mraw, INC ~ SPP * TRIAL, xlab= "Species.trial", ylab= "Number of errors")
boxplot(data=Mraw, TIME ~ SPP * TRIAL, xlab= "Species.trial", ylab= "Time to completion")


```
  
  
Now to the mixed models.

```{r, warning=FALSE, message=FALSE}
library(lme4)
library(lmerTest)
library(car)


m1 <- lmer(INC ~ TRIAL * SPP + (1|X.U.FEFF.NAME), data = Mraw)
summary(m1)
Anova(m1, type = 3)
Anova(m1, type=3, test.statistic = "F")

m2 <- lmer(TIME~TRIAL*SPP + (1|X.U.FEFF.NAME), data = Mraw)
summary(m2)
Anova(m2, type = 3)
Anova(m2, type=3, test.statistic = "F")


```

<font color= "deeporchid4"> Both model fits give me a warning about singular fit. I am unsure what this means or what to do about it. Not sure if I am coding the fixed and random effects correctly (should subjects be numbers rather than names? should there be a factor instead of 1 prior to random effect?). Think I want to understand this prior to trying the GLMM.</font>




#### **Second try** 

I changed the csv file to include subject number instead of name. Additionally, I will break the model down into 3 questions.

```{r, warning=FALSE, message=FALSE}
library(curl)

raw2 <- curl("https://raw.githubusercontent.com/allisondavis/complex_maze/master/mixed.format_complex-raw2.csv")

raw2 <- read.csv(raw2, header = TRUE, sep = ",", stringsAsFactors = TRUE)

raw2

names(raw2)[1] <- "SUBJECT"   #easier to type without the X.U.FEFF in front!

raw2

```




* **Model 1**: How does species affect errors/duration? Fixed effects: INC or TIME, SPP; Random effects: TRIAL, SUBJECT
  
```{r, warning=FALSE, message=FALSE}
library(lme4)
library(lmerTest)
library(car)


M1.e <- lmer(INC~SPP + (1|TRIAL)+(1|SUBJECT), data = raw2)
summary(M1.e)
Anova(M1.e, type = 3, test.statistic = "F")

M1.t <- lmer(TIME~SPP + (1|TRIAL)+(1|SUBJECT), data = raw2)
summary(M1.t)
Anova(M1.t, type = 3, test.statistic = "F")

```

<font color= "deeporchid4"> ***RESULTS:***</font> Even with subject as numbers, still getting a "boundary singular fit" error, with no variance in subject. Maybe sample size is too small? Maybe it has some variance, but it's so small that R doesn't report it?

Still get no significant result: there is no difference in errors or duration between species. In agreement with t-tests.



  
* **Model 2**: How does trial affect errors/duration? Aka was there any learning across trials? Fixed effects: INC or TIME, TRIAL; Random effects: SPP, SUBJECT
  
```{r, warning=FALSE, message=FALSE}
M2.e <- lmer(INC~TRIAL + (1|SPP)+(1|SUBJECT), data = raw2)
summary(M2.e)
Anova(M2.e, type = 3, test.statistic = "F")

M2.t <- lmer(TIME~TRIAL + (1|SPP)+(1|SUBJECT), data = raw2)
summary(M2.t)
Anova(M2.t, type = 3, test.statistic = "F")



library(multcomp)
library(lsmeans)

#summary(glht(M2.t, lsm(pairwise~TRIAL)))

```

<font color= "deeporchid4"> ***RESULTS:***</font> still get the singular fit error, this time with no variance in subject OR species. Error model shows no significance, which is in agreement with t-tests. Duration shows significance, which agrees with t-tests, but the summary of the lmm shows a different p-value than the Anova with the F statistic (lmm: p=0.0427, Anova: p= 0.05501). Not sure what to make of the difference or which one I should use. Additionally, it doesn't seem like this test will show me which trials are differing. Possibly another function will extract that (tried using multcomp and lsmeans packages, but still not sure how to make it work, keep getting errors).



  
* **Model 3**: How does trial and species affect errors/duration? Aka are there differences in learning between species? Fixed effects: INC or TIME, TRIAL, SPP; Random effects: SUBJECT


```{r}

M3.e <- lmer(INC~TRIAL*SPP + (1|SUBJECT), data = raw2)
summary(M3.e)
Anova(M3.e, type = 3, test.statistic = "F")

M3.e <- lmer(INC~SPP*TRIAL + (1|SUBJECT), data = raw2)   #just to see if the order makes a difference
summary(M3.e)
Anova(M3.e, type = 3, test.statistic = "F")

#order does not make a difference

M3.t <- lmer(TIME~TRIAL*SPP + (1|SUBJECT), data = raw2)
summary(M3.t)
Anova(M3.t, type = 3, test.statistic = "F")


```

<font color= "deeporchid4"> ***RESULTS:***</font> again with the singular fit and zero variance for subject. There are no significant results for either error or duration. I suppose this indicates that there were no differences in learning between species, but with trials significant, maybe this means that learning occured within species but no difference between? T-tests showed significant differences in duration for LAT T1-T2, and T1-T3, and FORM T2-T3, but again no differences between species for single trials. Maybe this is confirming that? 






  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  