---
title: "complex-maze"
author: "Allison Davis"
date: "February 15, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data collection

```{r, warning=FALSE, message=FALSE}
library(curl)

raw <- curl("https://raw.githubusercontent.com/allisondavis/complex_maze/master/adavis_complex-complete.csv")

raw <- read.csv(raw, header = TRUE, sep = ",", stringsAsFactors = TRUE)

head(raw)

```


## Analyses {.tabset .tabset-fade .tabset-pills}

###  First look

I will first look at the data via box plots for each trial vs species, as well as bar charts per species

```{r, warning=FALSE, message=FALSE}
library(ggplot2)

ggplot(raw, aes(x=SPP, y=T1.TIME, fill=SPP)) + geom_boxplot() + xlab("Species") + ylab("Time to completion (s)\n") + ggtitle("Trial 1\n") + geom_jitter(shape=16, position=position_jitter(0.2)) + theme_classic() + scale_fill_manual(values = c("#9C964A", "#FAD77B")) + theme(text = element_text(size = 16)) + theme(legend.position = "none")

ggplot(raw, aes(x=SPP, y=T2.TIME, fill=SPP)) + geom_boxplot() + xlab("Species") + ylab("Time to completion (s)\n") + ggtitle("Trial 2\n") + geom_jitter(shape=16, position=position_jitter(0.2)) + theme_classic() + scale_fill_manual(values = c("#9C964A", "#FAD77B")) + theme(text = element_text(size = 16)) + theme(legend.position = "none")

ggplot(raw, aes(x=SPP, y=T3.TIME, fill=SPP)) + geom_boxplot() + xlab("Species") + ylab("Time to completion (s)\n") + ggtitle("Trial 3\n") + geom_jitter(shape=16, position=position_jitter(0.2)) + theme_classic() + scale_fill_manual(values = c("#9C964A", "#FAD77B")) + theme(text = element_text(size = 16)) + theme(legend.position = "none")

#come back and get bar chart/boxplot/scatterplot to compare three trials within each spp

lat <- raw[raw$SPP == "lat",]
form <- raw[raw$SPP == "form",]

trial.er.lat <- lat[c(4,6,8)]
trial.tm.lat <- lat[9:11]

trial.er.form <- form[c(4,6,8)]
trial.tm.form <- form[c(9:11)]

par(mfrow=c(2,2))
boxplot(trial.er.lat, main="Errors for sailfin")
boxplot(trial.er.form, main="Errors for amazon")
boxplot(trial.tm.lat, main="Duration for sailfin")
boxplot(trial.tm.form, main ="Duration for amazon")


```


### Simple stats

* using t-test to detect differences in means between species in same trial, and between trials in same species

```{r}
#testing differences between species w/in same trial
t.test(lat$T1.I, form$T1.I, alternative = "two.sided")

t.test(lat$T2.I, form$T2.I, alternative = "two.sided")

t.test(lat$T3.I, form$T3.I, alternative = "two.sided")

#testing differences between trials w/in same species
t.test(lat$T1.I, lat$T2.I, alternative = "two.sided")

t.test(lat$T1.I, lat$T3.I, alternative = "two.sided")

t.test(lat$T2.I, lat$T3.I, alternative = "two.sided")


t.test(form$T1.I, form$T2.I, alternative = "two.sided")

t.test(form$T1.I, form$T3.I, alternative = "two.sided")

t.test(form$T2.I, form$T3.I, alternative = "two.sided")

```

<font color = "deeporchid4"> No significant differences found.</font>

* using F-test to detect different variations in different species in same trial, and different trials in same species

```{r}


#testing variances between species w/in same trial
var.test(lat$T1.I, form$T1.I, alternative = "two.sided")

var.test(lat$T2.I, form$T2.I, alternative = "two.sided")

var.test(lat$T3.I, form$T3.I, alternative = "two.sided")

#testing variances between trials w/in same species
var.test(lat$T1.I, lat$T2.I, alternative = "two.sided")

var.test(lat$T1.I, lat$T3.I, alternative = "two.sided")

var.test(lat$T2.I, lat$T3.I, alternative = "two.sided")


var.test(form$T1.I, form$T2.I, alternative = "two.sided")

var.test(form$T1.I, form$T3.I, alternative = "two.sided")

var.test(form$T2.I, form$T3.I, alternative = "two.sided")


```

* ***will repeat for time of trials***

* using t-test to detect differences in means between species in same trial, and between trials in same species

```{r}
#testing differences between species w/in same trial
t.test(lat$T1.TIME, form$T1.TIME, alternative = "two.sided")

t.test(lat$T2.TIME, form$T2.TIME, alternative = "two.sided")

t.test(lat$T3.TIME, form$T3.TIME, alternative = "two.sided")

#testing differences between trials w/in same species
t.test(lat$T1.TIME, lat$T2.TIME, alternative = "two.sided")

t.test(lat$T1.TIME, lat$T3.TIME, alternative = "two.sided")

t.test(lat$T2.TIME, lat$T3.TIME, alternative = "two.sided")


t.test(form$T1.TIME, form$T2.TIME, alternative = "two.sided")

t.test(form$T1.TIME, form$T3.TIME, alternative = "two.sided")

t.test(form$T2.TIME, form$T3.TIME, alternative = "two.sided")

```

<font color = "deeporchid4"> No significant differences found.</font>

* using F-test to detect different variations in different species in same trial, and different trials in same species

```{r}


#testing variances between species w/in same trial
var.test(lat$T1.TIME, form$T1.TIME, alternative = "two.sided")

var.test(lat$T2.TIME, form$T2.TIME, alternative = "two.sided")

var.test(lat$T3.TIME, form$T3.TIME, alternative = "two.sided")

#testing variances between trials w/in same species
var.test(lat$T1.TIME, lat$T2.TIME, alternative = "two.sided")

var.test(lat$T1.TIME, lat$T3.TIME, alternative = "two.sided")

var.test(lat$T2.TIME, lat$T3.TIME, alternative = "two.sided")


var.test(form$T1.TIME, form$T2.TIME, alternative = "two.sided")

var.test(form$T1.TIME, form$T3.TIME, alternative = "two.sided")

var.test(form$T2.TIME, form$T3.TIME, alternative = "two.sided")


```


<font color="deeporchid4"> Only significant difference is between trials 1 and 3 in amazons.</font>


###  Complex stats {.tabset .tabset-fade .tabset-pills}

#### **First try** 

Here I will use a generalized linear mixed-effects model to assess differences in the number of errors made across trials between the species.
  * species as fixed effect, individual ID as random effect, and trial number as covariate

Had to rearrange the data, so that each trial for each individual was on a separate line. Will upload now.

```{r, warning=FALSE, message=FALSE}
library(curl)

Mraw <- curl("https://raw.githubusercontent.com/allisondavis/complex_maze/master/mixed.format_complex-raw.csv")

Mraw <- read.csv(Mraw, header = TRUE, sep = ",", stringsAsFactors = TRUE)

head(Mraw)

```

Now let's check how the data are distributed. Normal distributions can use linear mixed models, whereas non-normal distributions can use generalized linear mixed models.

  
```{r, warning=FALSE, message=FALSE}
library(car)

qqp(Mraw$INC, "norm")  #fits normal fine, so can used LMM
qqp(Mraw$TIME, "norm") #fits ok, little off in top right; can try LMM but may want to try a different distribution
qqp(Mraw$TIME, "lnorm") #fits better, since no points are outside dashed lines; try out GLMM

boxplot(data=Mraw, INC ~ SPP * TRIAL, xlab= "Species.trial", ylab= "Number of errors")
boxplot(data=Mraw, TIME ~ SPP * TRIAL, xlab= "Species.trial", ylab= "Time to completion")


```
  
  
Now to the mixed models.

```{r, warning=FALSE, message=FALSE}
library(lme4)
library(lmerTest)
library(car)


m1 <- lmer(INC ~ TRIAL * SPP + (1|X.U.FEFF.NAME), data = Mraw)
summary(m1)
Anova(m1, type = 3)
Anova(m1, type=3, test.statistic = "F")

m2 <- lmer(TIME~TRIAL*SPP + (1|X.U.FEFF.NAME), data = Mraw)
summary(m2)
Anova(m2, type = 3)
Anova(m2, type=3, test.statistic = "F")


```

<font color= "deeporchid4"> Both model fits give me a warning about singular fit. I am unsure what this means or what to do about it. Not sure if I am coding the fixed and random effects correctly (should subjects be numbers rather than names? should there be a factor instead of 1 prior to random effect?). Think I want to understand this prior to trying the GLMM.</font>




#### **Second try** 

I changed the csv file to include subject number instead of name. Additionally, I will break the model down into 3 questions.

```{r, warning=FALSE, message=FALSE}
library(curl)

raw2 <- curl("https://raw.githubusercontent.com/allisondavis/complex_maze/master/mixed.format_complex-raw2.csv")

raw2 <- read.csv(raw2, header = TRUE, sep = ",", stringsAsFactors = TRUE)

raw2

names(raw2)[1] <- "SUBJECT"   #easier to type without the X.U.FEFF in front!

raw2

```




* **Model 1**: How does species affect errors/duration? Fixed effects: INC or TIME, SPP; Random effects: TRIAL, SUBJECT
  
```{r, warning=FALSE, message=FALSE}
library(lme4)
library(lmerTest)
library(car)


M1.e <- lmer(INC~SPP + (1|TRIAL)+(1|SUBJECT), data = raw2)
summary(M1.e)
Anova(M1.e, type = 3, test.statistic = "F")

M1.t <- lmer(TIME~SPP + (1|TRIAL)+(1|SUBJECT), data = raw2)
summary(M1.t)
Anova(M1.t, type = 3, test.statistic = "F")

```

<font color= "deeporchid4"> ***RESULTS:***</font> Even with subject as numbers, still getting a "boundary singular fit" error, with no variance in subject. Maybe sample size is too small? Maybe it has some variance, but it's so small that R doesn't report it?

Still get no significant result: there is no difference in errors or duration between species. In agreement with t-tests.



  
* **Model 2**: How does trial affect errors/duration? Aka was there any learning across trials? Fixed effects: INC or TIME, TRIAL; Random effects: SPP, SUBJECT
  
```{r, warning=FALSE, message=FALSE}
M2.e <- lmer(INC~TRIAL + (1|SPP)+(1|SUBJECT), data = raw2)
summary(M2.e)
Anova(M2.e, type = 3, test.statistic = "F")

M2.t <- lmer(TIME~TRIAL + (1|SPP)+(1|SUBJECT), data = raw2)
summary(M2.t)
Anova(M2.t, type = 3, test.statistic = "F")



library(multcomp)
library(lsmeans)

#summary(glht(M2.t, lsm(pairwise~TRIAL)))

```

<font color= "deeporchid4"> ***RESULTS:***</font> still get the singular fit error, this time with no variance in subject OR species. Error model shows no significance, which is in agreement with t-tests. Duration shows significance, which agrees with t-tests, but the summary of the lmm shows a different p-value than the Anova with the F statistic (lmm: p=0.0427, Anova: p= 0.05501). Not sure what to make of the difference or which one I should use. Additionally, it doesn't seem like this test will show me which trials are differing. Possibly another function will extract that (tried using multcomp and lsmeans packages, but still not sure how to make it work, keep getting errors).



  
* **Model 3**: How does trial and species affect errors/duration? Aka are there differences in learning between species? Fixed effects: INC or TIME, TRIAL, SPP; Random effects: SUBJECT


```{r}

M3.e <- lmer(INC~TRIAL*SPP + (1|SUBJECT), data = raw2)
summary(M3.e)
Anova(M3.e, type = 3, test.statistic = "F")

M3.e <- lmer(INC~SPP*TRIAL + (1|SUBJECT), data = raw2)   #just to see if the order makes a difference
summary(M3.e)
Anova(M3.e, type = 3, test.statistic = "F")

#order does not make a difference

M3.t <- lmer(TIME~TRIAL*SPP + (1|SUBJECT), data = raw2)
summary(M3.t)
Anova(M3.t, type = 3, test.statistic = "F")


```

<font color= "deeporchid4"> ***RESULTS:***</font> again with the singular fit and zero variance for subject. There are no significant results for either error or duration. I suppose this indicates that there were no differences in learning between species, but with trials significant, maybe this means that learning occurred within species but no difference between? T-tests showed significant differences in duration for LAT T1-T2, and T1-T3, and FORM T2-T3, but again no differences between species for single trials. Maybe this is confirming that? 



#### **Take 3**

Using same data set as before (where each trial is separate line).

```{r, warning=FALSE, message=FALSE}
library(curl)

raw2 <- curl("https://raw.githubusercontent.com/allisondavis/complex_maze/master/mixed.format_complex-raw2.csv")

raw2 <- read.csv(raw2, header = TRUE, sep = ",", stringsAsFactors = TRUE)

raw2

names(raw2)[1] <- "SUBJECT"   #easier to type without the X.U.FEFF in front!

raw2

```


***FOCUS***: the focus will be on model 3, where we see how learning across three trials differs between the species 
`[INC or TIME~TRIAL* SPP +(1|SUBJECT)])` 


  
```{r, warning=FALSE, message=FALSE}
library(lme4)
library(lmerTest)
library(car)

M3.e <- lmer(INC~TRIAL*SPP + (1|SUBJECT), data = raw2)
summary(M3.e)
Anova(M3.e, type = 3, test.statistic = "F")


M3.t <- lmer(TIME~TRIAL*SPP + (1|SUBJECT), data = raw2)
summary(M3.t)
Anova(M3.t, type = 3, test.statistic = "F")


```  
  


  * As we saw previously, we get a singular fit, zero variance and SD with SUBJECT as our random effect. Upon discussing this with the lab, they said it could indicate two things: either the effect of subject is negligent on the model, so we should run a model without subject, OR this is a byproduct of having too many variables to explain the variation. 


I will run a linear model without the effect of subject to see how it compares to the linear mixed model.


```{r}
LM3.e <- lm(INC~TRIAL*SPP, data = raw2)
summary(LM3.e)
Anova(LM3.e, type = 3, test.statistic = "F")


LM3.t <- lm(TIME~TRIAL*SPP, data = raw2)
summary(LM3.t)
Anova(LM3.t, type = 3, test.statistic = "F")



```
  

<font color= "deeporchid4"> ***RESULTS:***</font> We see very similar, if not identical p-values for the linear model. This seems to indicate that the mixed model is either not correctly taking into account the effect of subject, or subject really has no overall affect on INC or TIME. 




Another suggestion made was to run a *generalized* mixed linear model using the poisson distribution. This was suggested due to the small sample size, and the deviance from normalcy seen in our qqplots (see second try). I will try a generalized mixed linear model on the data with a poisson fit to see how this affects our results.

  * *Need to figure out how to do a qqplot with a poisson fit. Previous website I was using no longer exists!* (https://ase.tufts.edu/gsc/gradresources/guidetomixedmodelsinr/mixed%20model%20guide.html)




```{r, warning=FALSE, message=FALSE}

library(lme4)
library(lmerTest)
library(car)


GLM3.e <- glmer(INC~TRIAL*SPP + (1|SUBJECT), data = raw2, family="poisson")
summary(GLM3.e)


GLM3.t <- glmer(TIME~TRIAL*SPP + (1|SUBJECT), data = raw2, family="poisson")
summary(GLM3.t)



```
  
  
<font color= "deeporchid4"> ***RESULTS:*** </font> It appears that with INC, trial is now significant but species and interaction still is not. This would suggest that this is a change occuring across trials, but that there is no difference between the species (seems about right with boxplots). As for TIME, the summary throws a continuous error since all the time is non-integer (poisson runs with integers), but even when the error ends, we get trial and interaction of trial and species as significant (suuuper sig). Not really sure what it means to have interaction but not spp by itself as significant. Also, the Anova function didn't work for this, not sure why and not sure if we need and F statistic p-value or if the p-value given is fine.
  
  
  
  
  
  
  
  
  
  